---
title: "Figure 3"
author: "Amanda Welch"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gtools)
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(data.table)
library(ggplot2)
```

<!--
Figure 3: trial/ITI duration (3A/3B) and DLC nose-tracking speed (3C/3D),
DOI vs. saline, paired by mouse. 3C/3D depend on a separately-run DLC
tracking pipeline (see "dlc pipeline reference" chunk below, eval=FALSE --
included for reference/reproducibility, not executed by this notebook;
its output is loaded pre-computed from D:/R data/TrackingAll.Rds).
-->

```{r shared setup}
# mouse id, doi session number, and saline session number for each animal.
# used throughout 3a-3d, and reshaped into a lookup list for 3c/3d below.
mouse_sessions <- data.frame(
  mouse_id = c(2174, 2181, 2182, 2183, 2185, 2191, 2195, 2196, 2197, 2198, 2200, 2201, 2202, 2203, 2204, 2205, 2208, 2209, 2210, 2211, 2212, 2213),
  session_doi = c(57, 42, 32, 34, 41, 25, 17, 18, 17, 17, 14, 16, 19, 13, 17, 12, 18, 14, 17, 13, 16, 14),
  session_saline = c(56, 41, 31, 33, 40, 21, 20, 16, 19, 19, 19, 21, 14, 18, 22, 16, 14, 18, 13, 17, 13, 18)
)

# female mouse ids; all others are male
# 2174 f, 2181 f, 2182 m, 2183 m, 2185 m, 2191 f, 2195-98 m, 2200-2204 f,
# 2205-06 m, 2208-09 m, 2210-13 f
female_ids <- c(2174, 2181, 2191, 2200, 2201, 2202, 2203, 2204, 2210, 2211, 2212, 2213)

# loads every trialsummary.txt under root_directory, keeps only the 100-0
# condition, drops sessions with < 40 trials or a known N2/odor delivery
# error, and computes trial_length and iti_length for each remaining
# session. returns a named list of data frames (name = source file path).
load_trial_summaries <- function(root_directory = "D:/newmove") {
  trial_summ <- list.files(path = root_directory,
                           recursive = TRUE,
                           pattern = "trialsummary.txt",
                           full.names = TRUE)

  # 100-0 condition only, sorted numerically so sessions import in order
  trial_summ_sorted <- mixedsort(trial_summ[grep("100-0", trial_summ)])

  # mouse 2191 session 20 is a blank file (no trials completed); it reads
  # in as a 0-row data frame and gets dropped by the <40-trial filter below
  all_data <- lapply(trial_summ_sorted, read.table, header = FALSE, sep = ",")
  names(all_data) <- trial_summ_sorted

  data_above40trials <- purrr::discard(all_data, ~nrow(.) < 40)

  col_names <- c("trial", "odor_setting", "odor_side", "success", "tstart", "tend", "set_iti")
  for (i in seq_along(data_above40trials)) {
    colnames(data_above40trials[[i]]) <- col_names
  }

  # manually discard sessions with n2/odor delivery errors, per experimenter notes
  # 2174 session 34: error on left valve - fixed after session
  # 2174 session 24: n2 not turned on for half of session
  # 2174 session 92: left n2 had problems
  # 2185 session 74: left n2 had problems, same day as 2174 session 92
  excluded <- c(
    "D:Mouse_2174/iti_intervals/34/trialsummary.txt",
    "D:Mouse_2174/iti_intervals/24/trialsummary.txt",
    "D:Mouse_2174/iti_intervals/92/trialsummary.txt",
    "D:Mouse_2185/iti_intervals/74/trialsummary.txt"
  )
  data_above40trials <- data_above40trials[!names(data_above40trials) %in% excluded]

  data_above40trials <- lapply(data_above40trials, transform, trial_length = tend - tstart)
  data_above40trials <- lapply(data_above40trials, transform, iti_length = tstart - lag(tend))

  # shift iti_length up one row: as computed above, iti_length[i] is the gap
  # before trial i, but it belongs with the trial that follows the gap
  shift <- function(x, n) c(x[-(seq(n))], rep(NA, n))
  for (i in seq_along(data_above40trials)) {
    data_above40trials[[i]]$iti_length <- shift(data_above40trials[[i]]$iti_length, 1)
  }

  data_above40trials
}
```

# Figure 3A

trial duration on odor trials, DOI vs. saline, paired by mouse.

```{r 3a build stats}
data_above40trials <- load_trial_summaries()

# per-session summary stats, one row per session
full_stats <- data_above40trials %>%
  map(~summarise(group_by(., odor_setting),
                 trials_performed = length(odor_setting),
                 avg_trial_length = mean(trial_length),
                 sd_trial_length = sd(trial_length),
                 percent_correct = mean(success) * 100))

# collapse the nested list into one data frame, with mouse id and session
# number pulled from the source file path
stats_list <- setDT(do.call(rbind.data.frame, full_stats), keep.rownames = TRUE)[]
stats <- stats_list %>%
  mutate(
    id = as.numeric(sub('.*/([0-9]+)/100-0/.*', '\\1', rn)),
    session = as.numeric(sub('.*/100-0/([0-9]+)/.*', '\\1', rn))
  ) %>%
  select(-rn)

# tag each session as doi or saline using the lookup table, and drop any
# session that isn't one of the two designated dosing sessions per mouse
stats <- stats %>%
  mutate(
    is_doi = mapply(function(i, s) any(mouse_sessions$mouse_id == i & mouse_sessions$session_doi == s), id, session),
    is_saline = mapply(function(i, s) any(mouse_sessions$mouse_id == i & mouse_sessions$session_saline == s), id, session),
    sex = ifelse(id %in% female_ids, "F", "M")
  )

# odor trials (odor_setting == 0) for the main 3a comparison, and omission
# trials (odor_setting == 3) for the companion comparison reported in the
# text alongside 3a
odor_trials <- stats %>% filter(odor_setting == 0, is_doi | is_saline) %>%
  mutate(condition = ifelse(is_doi, "DOI", "Saline"))

omission_trials <- stats %>% filter(odor_setting == 3, is_doi | is_saline) %>%
  mutate(condition = ifelse(is_doi, "DOI", "Saline"))
```

```{r 3a build plot data}
paired_data_wide_trialdur <- odor_trials %>%
  select(id, avg_trial_length, condition, sex) %>%
  tidyr::pivot_wider(
    names_from = condition,
    values_from = avg_trial_length,
    names_prefix = "avg_trial_"
  )

group_stats_3a <- paired_data_wide_trialdur %>%
  group_by(sex) %>%
  summarize(
    doi_mean = mean(avg_trial_DOI, na.rm = TRUE),
    doi_median = median(avg_trial_DOI, na.rm = TRUE),
    saline_mean = mean(avg_trial_Saline, na.rm = TRUE),
    saline_median = median(avg_trial_Saline, na.rm = TRUE),
    n = n()
  )
```

```{r 3a plot}
paired_data_wide_trialdur %>%
  ggplot(aes(x = avg_trial_Saline,
             y = avg_trial_DOI,
             shape = sex,
             color = sex)) +
  geom_point(size = 6,
             stroke = 1.5,
             alpha = 0.8) +

  # add group medians
  geom_point(data = group_stats_3a,
             aes(x = saline_median,
                 y = doi_median,
                 color = sex),
             size = 10,
             shape = 1,   # big o
             stroke = 2.5,
             inherit.aes = FALSE) +

  scale_color_manual(values = c("M" = "#00B19E",
                                "F" = "#990068"),
                     guide = "none") +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dotted",
              color = "black",
              size = 1) +
  labs(title = "Odor trial duration",
       x = "Saline duration (s)",
       y = "DOI duration (s)",
       shape = "Sex") +
  coord_fixed(ratio = 1) +
  scale_x_continuous(limits = c(1.4, 6)) +
  scale_y_continuous(limits = c(1.4, 6)) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.15),
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2)) +
  guides(
    shape = guide_legend(override.aes = list(color = c("#990068", "#00B19E")))  # merge shape & color
  )
```

```{r 3a wilcoxon}
wilcox_3a <- wilcox.test(paired_data_wide_trialdur$avg_trial_Saline,
                         paired_data_wide_trialdur$avg_trial_DOI,
                         paired = TRUE)
print(wilcox_3a)
```

trial duration on omission trials, DOI vs. saline - reported alongside
figure 3a in the text but not plotted as its own panel.

```{r omission trial duration wilcoxon}
paired_data_wide_omission <- omission_trials %>%
  select(id, avg_trial_length, condition, sex) %>%
  tidyr::pivot_wider(
    names_from = condition,
    values_from = avg_trial_length,
    names_prefix = "avg_trial_"
  )

wilcox_omission <- wilcox.test(paired_data_wide_omission$avg_trial_Saline,
                               paired_data_wide_omission$avg_trial_DOI,
                               paired = TRUE)
print(wilcox_omission)
```

# Figure 3B

ITI duration, DOI vs. saline, paired by mouse.

```{r 3b build data}
data_above40trials <- load_trial_summaries()

# split sessions into doi and saline lists by matching each file path's
# mouse id / session number against mouse_sessions
extract_id_session <- function(name) {
  matches <- str_match(name, ".*/(\\d+)/.*/(\\d+)/trialsummary\\.txt$")
  if (is.na(matches[1])) {
    return(list(id = NA, session = NA))
  }
  list(id = as.numeric(matches[2]), session = as.numeric(matches[3]))
}

filter_sessions <- function(data_list, session_col) {
  data_list[
    sapply(names(data_list), function(name) {
      parsed <- extract_id_session(name)
      if (is.na(parsed$id)) {
        return(FALSE)
      }
      any(mouse_sessions$mouse_id == parsed$id & mouse_sessions[[session_col]] == parsed$session)
    })
  ]
}

doi_dfs <- filter_sessions(data_above40trials, "session_doi")
saline_dfs <- filter_sessions(data_above40trials, "session_saline")

# per-mouse summary stats for each condition
summarize_trials <- function(df, mouse_id) {
  df %>%
    summarise(
      id = mouse_id,
      trials_performed = n(),
      avg_trial_length = mean(trial_length, na.rm = TRUE),
      sd_trial_length = sd(trial_length, na.rm = TRUE),
      avg_iti_length = mean(iti_length, na.rm = TRUE),
      sd_iti_length = sd(iti_length, na.rm = TRUE),
      percent_correct = mean(success, na.rm = TRUE) * 100,
      .groups = "drop"
    )
}

doi_stats <- map2_df(doi_dfs, names(doi_dfs), ~summarize_trials(.x, extract_id_session(.y)$id))
saline_stats <- map2_df(saline_dfs, names(saline_dfs), ~summarize_trials(.x, extract_id_session(.y)$id))

combined_stats <- full_join(doi_stats, saline_stats, by = "id", suffix = c("_doi", "_saline")) %>%
  mutate(sex = ifelse(id %in% female_ids, "F", "M"))

group_stats_3b <- combined_stats %>%
  group_by(sex) %>%
  summarize(
    doi_mean = mean(avg_iti_length_doi, na.rm = TRUE),
    doi_median = median(avg_iti_length_doi, na.rm = TRUE),
    saline_mean = mean(avg_iti_length_saline, na.rm = TRUE),
    saline_median = median(avg_iti_length_saline, na.rm = TRUE),
    n = n()
  )
```

```{r 3b plot}
combined_stats %>%
  ggplot(aes(x = avg_iti_length_saline,
             y = avg_iti_length_doi,
             shape = sex,
             color = sex)) +
  geom_point(size = 6,
             stroke = 1.5,
             alpha = 0.8) +

  # add group medians
  geom_point(data = group_stats_3b,
             aes(x = saline_median,
                 y = doi_median,
                 color = sex),
             size = 10,
             shape = 1,   # big o
             stroke = 2.5,
             inherit.aes = FALSE) +

  scale_color_manual(values = c("M" = "#00B19E",
                                "F" = "#990068"),
                     guide = "none") +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dotted",
              color = "black",
              size = 1) +
  labs(title = "ITI duration",
       x = "Saline duration (s)",
       y = "DOI duration (s)",
       shape = "Sex") +
  coord_fixed(ratio = 1) +
  scale_x_continuous(limits = c(13, 50)) +
  scale_y_continuous(limits = c(13, 50)) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.15),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.title = element_blank(),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2)) +
  guides(
    shape = guide_legend(override.aes = list(color = c("#990068", "#00B19E")))  # merge shape & color
  )
```

```{r 3b wilcoxon}
wilcox_3b <- wilcox.test(combined_stats$avg_iti_length_saline,
                         combined_stats$avg_iti_length_doi,
                         paired = TRUE)
print(wilcox_3b)
```

# Figure 3C and 3D setup

speed during the task (DLC nose tracking), split into within-trial (3c) and
between-trial/ITI (3d) periods.

```{r include=FALSE}
library(sp)
library(imputeTS)
library(ggplot2)
library(ggmap)
library(cowplot)
library(corrplot)
library(keras)
```

```{r dlc pipeline reference, eval=FALSE}
source('R/DLCAnalyzer_Functions_final_custom.R')

CleanTracking <- function(t, likelihoodcutoff = 0.95, existence.pol = NULL, maxdelta = NULL) {
  if (!IsTrackingData(t)) stop("Object is not of type TrackingData")

  for (i in 1:length(t$data)) {
    process <- t$data[[i]]$likelihood < likelihoodcutoff

    if (!is.null(existence.pol)) {
      process <- process | !point.in.polygon(t$data[[i]]$x, t$data[[i]]$y, existence.pol$x, existence.pol$y)
    }
    if (!is.null(maxdelta)) {
      dx <- c(NA, diff(t$data[[i]]$x))
      dy <- c(NA, diff(t$data[[i]]$y))
      delta <- sqrt(dx^2 + dy^2)
      process <- process | (delta > maxdelta)
    }

    t$data[[i]]$x[process] <- NA
    t$data[[i]]$y[process] <- NA
    t$data[[i]] <- na_interpolation(t$data[[i]])
  }

  return(t)
}

pipeline <- function(path) {
  Tracking <- ReadDLCDataFromCSV(path, fps = 80)
  Tracking <- CleanTracking(Tracking, likelihoodcutoff = 0.95, maxdelta = 8)
  Tracking <- OFTAnalysis(Tracking, movement_cutoff = 5, integration_period = 5, points = "Nose")
  return(Tracking)
}

RunPipeline_aw <- function(files, path, FUN) {
  out <- list()
  for (j in files) {
    out[[paste(j)]] <- FUN(file.path(path, j))
  }
  return(out)
}
```

```{r load tracking}
TrackingAll <- readRDS("D:/R data/TrackingAll.Rds")

# compute frame-to-frame speed (cm/s) from the nose tracking point
for (name in names(TrackingAll)) {
  df <- TrackingAll[[name]][["data"]][["Nose"]]

  delta_x <- c(NA, diff(df$x))
  delta_y <- c(NA, diff(df$y))
  delta_frame <- c(NA, diff(df$frame))  # should be 1 if no dropped frames

  df$speed_aw <- sqrt(delta_x^2 + delta_y^2) * 80 / delta_frame  # px/s (80 fps)
  df$speed_aw_cm <- df$speed_aw * 0.0237  # px-to-cm conversion factor

  TrackingAll[[name]][["data"]][["Nose"]] <- df
}
```

```{r split doi and saline tracking}
# pulls the nose-tracking data frame for one mouse/session/condition out of
# TrackingAll (named "<mouse_id>_<session>_<condition>.csv")
get_nose_tracking <- function(tracking_all, mouse_id, session, condition) {
  file_name <- paste0(mouse_id, "_", session, "_", condition, ".csv")
  tracking_data <- tracking_all[[file_name]]
  if (is.null(tracking_data) || is.null(tracking_data$data$Nose)) {
    return(NULL)
  }
  tracking_data$data$Nose
}

doi_tracking_data <- list()
saline_tracking_data <- list()

for (file_name in names(TrackingAll)) {
  parts <- strsplit(file_name, "_")[[1]]
  mouse_id <- parts[1]
  session <- parts[2]

  if (grepl("_doi\\.csv$", file_name, ignore.case = TRUE)) {
    doi_tracking_data[[file_name]] <- get_nose_tracking(TrackingAll, mouse_id, session, "doi")
  } else if (grepl("_saline\\.csv$", file_name, ignore.case = TRUE)) {
    saline_tracking_data[[file_name]] <- get_nose_tracking(TrackingAll, mouse_id, session, "saline")
  }
}
```

```{r trial timing lookup}
# ids/sessions to keep, in the "<mouse_id> = c(doi_session, saline_session)"
# form needed for file-path matching below (derived from mouse_sessions)
keep_ids_sessions <- setNames(
  lapply(seq_len(nrow(mouse_sessions)), function(i) {
    c(mouse_sessions$session_doi[i], mouse_sessions$session_saline[i])
  }),
  as.character(mouse_sessions$mouse_id)
)

is_matching_file <- function(path) {
  parts <- strsplit(path, "/")[[1]]
  id <- parts[3]
  session <- as.numeric(parts[5])
  id %in% names(keep_ids_sessions) && session %in% keep_ids_sessions[[id]]
}

# trial-level data (tstart/tend/odor_setting) for the doi/saline sessions
# only, used below to label each tracking frame as within- or
# between-trial
trial_summ <- list.files(path = "D:/newmove",
                         recursive = TRUE,
                         pattern = "trialsummary.txt",
                         full.names = TRUE)
trial_summ_iti <- trial_summ[grep("100-0", trial_summ)]
trial_summ_filtered <- Filter(is_matching_file, trial_summ_iti)
trial_summ_sorted <- mixedsort(trial_summ_filtered)

trial_data_lookup <- lapply(trial_summ_sorted, read.table, header = FALSE, sep = ",")
names(trial_data_lookup) <- trial_summ_sorted

col_names <- c("trial", "odor_setting", "odor_side", "success", "tstart", "tend", "set_iti")
for (i in seq_along(trial_data_lookup)) {
  colnames(trial_data_lookup[[i]]) <- col_names
}

trial_data_lookup <- lapply(trial_data_lookup, function(df) {
  transform(df, trial_length = tend - tstart, iti_length = tstart - dplyr::lag(tend))
})

shift <- function(x, n) c(x[-(seq(n))], rep(NA, n))
for (i in seq_along(trial_data_lookup)) {
  trial_data_lookup[[i]]$iti_length <- shift(trial_data_lookup[[i]]$iti_length, 1)
}
```

```{r label trial periods}
# labels each tracking frame with the trial period it falls in:
# "Odor" (odor_setting == 0), "Omission" (odor_setting == 3), or
# "Between trials" (default, outside any trial's tstart-tend window).
# trial_label_pooled collapses Odor + Omission into "Within trial", for
# the pooled-trial-type comparisons used in figures 3c/3d.
add_trial_label <- function(tracking_data, trial_data) {
  tracking_data$trial_label <- "Between trials"

  if (nrow(trial_data) > 0) {
    timestamps <- (1:nrow(tracking_data)) / 80  # frame index -> time (s)

    for (i in 1:nrow(trial_data)) {
      in_trial <- timestamps >= trial_data$tstart[i] & timestamps <= trial_data$tend[i]
      label <- if (trial_data$odor_setting[i] == 3) "Omission" else "Odor"
      tracking_data$trial_label[in_trial] <- label
    }
  }

  tracking_data$trial_label_pooled <- ifelse(tracking_data$trial_label %in% c("Odor", "Omission"),
                                             "Within trial", "Between trials")
  tracking_data
}

label_all_trial_periods <- function(tracking_list, trial_lookup) {
  for (i in seq_along(tracking_list)) {
    file_name <- names(tracking_list)[i]
    parts <- strsplit(file_name, "_")[[1]]
    mouse_id <- parts[1]
    session <- parts[2]

    trial_data_key <- paste0("D:/newmove/", mouse_id, "/100-0/", session, "/trialsummary.txt")

    if (trial_data_key %in% names(trial_lookup)) {
      tracking_list[[i]] <- add_trial_label(tracking_list[[i]], trial_lookup[[trial_data_key]])
    } else {
      message(paste("warning: no trial data for", trial_data_key))
      tracking_list[[i]]$trial_label <- "Between trials"
      tracking_list[[i]]$trial_label_pooled <- "Between trials"
    }
  }
  tracking_list
}

doi_tracking_data_processed <- label_all_trial_periods(doi_tracking_data, trial_data_lookup)
saline_tracking_data_processed <- label_all_trial_periods(saline_tracking_data, trial_data_lookup)
```

```{r build speed comparison}
mouse_ids <- substr(names(doi_tracking_data), 1, 4)

# per-mouse mean speed_aw_cm (excluding tracking artifacts above 160 cm/s),
# for each condition x trial-period combination. "Within"/"Between" here
# use the pooled (odor + omission combined) label, for figures 3c/3d.
speed_comparison <- data.frame(Mouse_ID = mouse_ids,
                               Mean_Speed_DOI_Within = NA,
                               Mean_Speed_DOI_Between = NA,
                               Mean_Speed_Saline_Within = NA,
                               Mean_Speed_Saline_Between = NA)

# per-mouse mean speed by detailed trial period (odor / omission / between),
# for the odor-vs-omission-vs-iti comparisons reported alongside 3c/3d
speed_comparison_detail <- data.frame(Mouse_ID = mouse_ids,
                                      Mean_Speed_DOI_Odor = NA,
                                      Mean_Speed_DOI_Omission = NA,
                                      Mean_Speed_DOI_Between = NA,
                                      Mean_Speed_Saline_Odor = NA,
                                      Mean_Speed_Saline_Omission = NA,
                                      Mean_Speed_Saline_Between = NA)

for (i in seq_along(mouse_ids)) {
  doi_filtered <- doi_tracking_data_processed[[i]][doi_tracking_data_processed[[i]]$speed_aw_cm <= 160, ]
  saline_filtered <- saline_tracking_data_processed[[i]][saline_tracking_data_processed[[i]]$speed_aw_cm <= 160, ]

  speed_comparison[i, "Mean_Speed_DOI_Within"] <- mean(doi_filtered[doi_filtered$trial_label_pooled == "Within trial", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison[i, "Mean_Speed_DOI_Between"] <- mean(doi_filtered[doi_filtered$trial_label_pooled == "Between trials", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison[i, "Mean_Speed_Saline_Within"] <- mean(saline_filtered[saline_filtered$trial_label_pooled == "Within trial", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison[i, "Mean_Speed_Saline_Between"] <- mean(saline_filtered[saline_filtered$trial_label_pooled == "Between trials", "speed_aw_cm"], na.rm = TRUE)

  speed_comparison_detail[i, "Mean_Speed_DOI_Odor"] <- mean(doi_filtered[doi_filtered$trial_label == "Odor", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison_detail[i, "Mean_Speed_DOI_Omission"] <- mean(doi_filtered[doi_filtered$trial_label == "Omission", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison_detail[i, "Mean_Speed_DOI_Between"] <- mean(doi_filtered[doi_filtered$trial_label == "Between trials", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison_detail[i, "Mean_Speed_Saline_Odor"] <- mean(saline_filtered[saline_filtered$trial_label == "Odor", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison_detail[i, "Mean_Speed_Saline_Omission"] <- mean(saline_filtered[saline_filtered$trial_label == "Omission", "speed_aw_cm"], na.rm = TRUE)
  speed_comparison_detail[i, "Mean_Speed_Saline_Between"] <- mean(saline_filtered[saline_filtered$trial_label == "Between trials", "speed_aw_cm"], na.rm = TRUE)
}

speed_comparison$sex <- ifelse(speed_comparison$Mouse_ID %in% female_ids, "F", "M")
speed_comparison_detail$sex <- ifelse(speed_comparison_detail$Mouse_ID %in% female_ids, "F", "M")

group_stats_speed <- speed_comparison %>%
  group_by(sex) %>%
  summarize(
    doi_within = median(Mean_Speed_DOI_Within, na.rm = TRUE),
    doi_between = median(Mean_Speed_DOI_Between, na.rm = TRUE),
    saline_within = median(Mean_Speed_Saline_Within, na.rm = TRUE),
    saline_between = median(Mean_Speed_Saline_Between, na.rm = TRUE),
    n = n()
  )
```

# Figure 3C

```{r 3c plot}
speed_comparison %>%
  ggplot(aes(x = Mean_Speed_Saline_Within,
             y = Mean_Speed_DOI_Within,
             color = sex,
             shape = sex)) +
  geom_point(size = 6,
             stroke = 1.5,
             alpha = 0.8) +

  # add group medians
  geom_point(data = group_stats_speed,
             aes(x = saline_within,
                 y = doi_within,
                 color = sex),
             size = 10,
             shape = 1,   # big o
             stroke = 2.5,
             inherit.aes = FALSE) +

  scale_color_manual(values = c("M" = "#00B19E",
                                "F" = "#990068"),
                     guide = "none") +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dotted",
              color = "black",
              size = 1) +
  labs(title = "Speed within trial",
       x = "Saline speed (cm/s)",
       y = "DOI speed (cm/s)",
       shape = "Sex") +
  scale_x_continuous(limits = c(3.5, 13)) +
  scale_y_continuous(limits = c(3.5, 13)) +
  coord_fixed(ratio = 1) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.15),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.title = element_blank(),
        text = element_text(size = 20),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2)) +
  guides(shape = guide_legend(override.aes = list(color = c("#990068", "#00B19E"))))
```

```{r 3c wilcoxon}
wilcox_3c <- wilcox.test(speed_comparison$Mean_Speed_DOI_Within,
                         speed_comparison$Mean_Speed_Saline_Within,
                         paired = TRUE)
print(wilcox_3c)
```

# Figure 3D

```{r 3d plot}
speed_comparison %>%
  ggplot(aes(x = Mean_Speed_Saline_Between,
             y = Mean_Speed_DOI_Between,
             color = sex,
             shape = sex)) +
  geom_point(size = 6,
             stroke = 1.5,
             alpha = 0.8) +

  # add group medians
  geom_point(data = group_stats_speed,
             aes(x = saline_between,
                 y = doi_between,
                 color = sex),
             size = 10,
             shape = 1,   # big o
             stroke = 2.5,
             inherit.aes = FALSE) +

  scale_color_manual(values = c("M" = "#00B19E",
                                "F" = "#990068"),
                     guide = "none") +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dotted",
              color = "black",
              size = 1) +
  labs(title = "Speed during ITI",
       x = "Saline speed (cm/s)",
       y = "DOI speed (cm/s)",
       shape = "Sex") +
  scale_x_continuous(limits = c(3.5, 13)) +
  scale_y_continuous(limits = c(3.5, 13)) +
  coord_fixed(ratio = 1) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.15),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.title = element_blank(),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2)) +
  guides(shape = guide_legend(override.aes = list(color = c("#990068", "#00B19E"))))
```

```{r 3d wilcoxon}
wilcox_3d <- wilcox.test(speed_comparison$Mean_Speed_DOI_Between,
                         speed_comparison$Mean_Speed_Saline_Between,
                         paired = TRUE)
print(wilcox_3d)
```

# Figure supplemental

```{r all pairwise speed comparisons}
library(gt)

speed_columns <- c("Mean_Speed_DOI_Odor", "Mean_Speed_DOI_Omission", "Mean_Speed_DOI_Between",
                   "Mean_Speed_Saline_Odor", "Mean_Speed_Saline_Omission", "Mean_Speed_Saline_Between")

# every pairwise combination of the six speed columns, each as a paired
# wilcoxon test (same mice measured under every condition/trial-period)
speed_comparisons <- combn(speed_columns, 2, simplify = FALSE)

speed_wilcox_results <- map(speed_comparisons, ~ wilcox.test(
  speed_comparison_detail[[.x[1]]],
  speed_comparison_detail[[.x[2]]],
  paired = TRUE
))
names(speed_wilcox_results) <- map_chr(speed_comparisons, ~ paste(.x, collapse = " vs "))

speed_wilcox_results
```

```{r speed comparisons summary table}
# tidy summary, sorted by p-value, with cleaned-up comparison labels and
# significance stars
speed_wilcox_summary <- map2_df(speed_comparisons, speed_wilcox_results, ~ tibble(
  Comparison = paste(.x, collapse = " vs "),
  V_statistic = .y$statistic,
  P_value = .y$p.value
)) %>%
  arrange(P_value) %>%
  mutate(
    Comparison = str_replace_all(Comparison, "Mean_Speed_", ""),
    Comparison = str_replace_all(Comparison, "Between", "ITI"),
    Comparison = str_replace_all(Comparison, "Odor", "odor"),
    Comparison = str_replace_all(Comparison, "Omission", "omission"),
    Comparison = str_replace_all(Comparison, "_", " "),
    Significance = case_when(
      P_value < 0.001 ~ "***",
      P_value < 0.01  ~ "**",
      P_value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )
  )

speed_table <- speed_wilcox_summary %>%
  gt() %>%
  fmt_number(columns = c(V_statistic, P_value), decimals = 4) %>%
  tab_header(
    title = "Speed comparisons"
  ) %>%
  cols_label(
    Comparison   = "Comparison",
    V_statistic  = "V",
    P_value      = "P-value",
    Significance = "Sig."
  ) %>%
  tab_options(
    table.font.size = "medium",
    heading.title.font.size = "large"
  )

speed_table
```
