---
title: "Figure 5"
author: "Amanda Welch"
output: html_document
---

this file produces the data (as csvs) and statistics for figure 5
(occupancy and head-twitch response location/rate). plotting is done
separately in occupancy_grandmean_gaussian.mlx (matlab) - this script
only writes the csvs that script reads in, plus runs the reported stats.
r and matlab code are kept in separate files intentionally.

<!--
Two things worth knowing before re-running this file:
- The `write.csv` calls for the occupancy_*.csv files that the MATLAB
  script needs (see "export occupancy csvs" chunk near the end) are
  commented out. A fresh run of this notebook will NOT regenerate those
  CSVs -- they only exist because someone ran that block manually at some
  point. Uncomment before relying on a from-scratch re-run.
- `bin_width` for the HTR grand-mean CSVs (below) is 100 px in this
  version of the file. If you find more than one copy of
  htr_grandmean_iti/htr_grandmean_within on the drive with different bin
  spacing (120 vs 100 px seen once already) and different values, this is
  almost certainly why: this script was re-run after `bin_width` changed,
  and the old output was never deleted.
-->

<!-- "fixes the broken export noted at the top of this file" (chunk
     "occupancy all 22 mice" below) doesn't refer to anything actually at
     the top of this file as it stands now -- likely a stale comment left
     over from an earlier version. Left as-is, not resolved. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(purrr)
library(stringr)
library(tidyr)
library(ggplot2)
library(ggbeeswarm)

# libraries used by the DLCAnalyzer package (sourced below)
library(sp)
library(imputeTS)
library(ggmap)
library(cowplot)
library(corrplot)
library(keras)
library(data.table)
```

# tracking, speed, and trial-period labeling

```{r shared setup}
mouse_sessions <- data.frame(
  mouse_id = c(2174, 2181, 2182, 2183, 2185, 2191, 2195, 2196, 2197, 2198, 2200, 2201, 2202, 2203, 2204, 2205, 2208, 2209, 2210, 2211, 2212, 2213),
  session_doi = c(57, 42, 32, 34, 41, 25, 17, 18, 17, 17, 14, 16, 19, 13, 17, 12, 18, 14, 17, 13, 16, 14),
  session_saline = c(56, 41, 31, 33, 40, 21, 20, 16, 19, 19, 19, 21, 14, 18, 22, 16, 14, 18, 13, 17, 13, 18)
)

# female mouse ids; all others are male
female_ids <- c(2174, 2181, 2191, 2200, 2201, 2202, 2203, 2204, 2210, 2211, 2212, 2213)

# 1st 5 mice (recorded on an earlier camera setup) are flipped on x-coord
flip_ids <- c("2174", "2181", "2182", "2183", "2185")

source("C:/Users/Amanda Welch/Desktop/Ph.D. Program/Lab/Code/DLCAnalyzer-master/R/DLCAnalyzer_Functions_final_custom.R")

TrackingAll <- readRDS("D:/R data/TrackingAll.Rds")

# compute frame-to-frame speed (cm/s) from the nose tracking point
for (name in names(TrackingAll)) {
  df <- TrackingAll[[name]][["data"]][["Nose"]]
  delta_x <- c(NA, diff(df$x))
  delta_y <- c(NA, diff(df$y))
  delta_frame <- c(NA, diff(df$frame))
  df$speed_aw <- sqrt(delta_x^2 + delta_y^2) * 80 / delta_frame
  df$speed_aw_cm <- df$speed_aw * 0.0237
  TrackingAll[[name]][["data"]][["Nose"]] <- df
}

get_nose_tracking <- function(tracking_all, mouse_id, session, condition) {
  file_name <- paste0(mouse_id, "_", session, "_", condition, ".csv")
  tracking_data <- tracking_all[[file_name]]
  if (is.null(tracking_data) || is.null(tracking_data$data$Nose)) return(NULL)
  tracking_data$data$Nose
}

doi_tracking_data <- list()
saline_tracking_data <- list()
for (file_name in names(TrackingAll)) {
  parts <- strsplit(file_name, "_")[[1]]
  mouse_id <- parts[1]
  session <- parts[2]
  if (grepl("_doi\\.csv$", file_name, ignore.case = TRUE)) {
    doi_tracking_data[[file_name]] <- get_nose_tracking(TrackingAll, mouse_id, session, "doi")
  } else if (grepl("_saline\\.csv$", file_name, ignore.case = TRUE)) {
    saline_tracking_data[[file_name]] <- get_nose_tracking(TrackingAll, mouse_id, session, "saline")
  }
}

# trial-level data (tstart/tend/odor_setting), used to label frames
keep_ids_sessions <- setNames(
  lapply(seq_len(nrow(mouse_sessions)), function(i) c(mouse_sessions$session_doi[i], mouse_sessions$session_saline[i])),
  as.character(mouse_sessions$mouse_id)
)
is_matching_file <- function(path) {
  parts <- strsplit(path, "/")[[1]]
  id <- parts[3]
  session <- as.numeric(parts[5])
  id %in% names(keep_ids_sessions) && session %in% keep_ids_sessions[[id]]
}

trial_summ <- list.files(path = "D:/newmove", recursive = TRUE, pattern = "trialsummary.txt", full.names = TRUE)
trial_summ_iti <- trial_summ[grep("100-0", trial_summ)]
trial_summ_filtered <- Filter(is_matching_file, trial_summ_iti)
trial_summ_sorted <- gtools::mixedsort(trial_summ_filtered)

trial_data_lookup <- lapply(trial_summ_sorted, read.table, header = FALSE, sep = ",")
names(trial_data_lookup) <- trial_summ_sorted
trial_col_names <- c("trial", "odor_setting", "odor_side", "success", "tstart", "tend", "set_iti")
for (i in seq_along(trial_data_lookup)) colnames(trial_data_lookup[[i]]) <- trial_col_names
trial_data_lookup <- lapply(trial_data_lookup, function(df) {
  transform(df, trial_length = tend - tstart, iti_length = tstart - dplyr::lag(tend))
})
shift <- function(x, n) c(x[-(seq(n))], rep(NA, n))
for (i in seq_along(trial_data_lookup)) trial_data_lookup[[i]]$iti_length <- shift(trial_data_lookup[[i]]$iti_length, 1)

# labels each tracking frame "Within trial" or "Between trials"
add_trial_label <- function(tracking_data, trial_data) {
  tracking_data$trial_label <- "Between trials"
  if (nrow(trial_data) > 0) {
    timestamps <- (1:nrow(tracking_data)) / 80
    for (i in 1:nrow(trial_data)) {
      in_trial <- timestamps >= trial_data$tstart[i] & timestamps <= trial_data$tend[i]
      tracking_data$trial_label[in_trial] <- "Within trial"
    }
  }
  tracking_data
}

label_all_trial_periods <- function(tracking_list, trial_lookup) {
  for (i in seq_along(tracking_list)) {
    file_name <- names(tracking_list)[i]
    parts <- strsplit(file_name, "_")[[1]]
    mouse_id <- parts[1]
    session <- parts[2]
    trial_data_key <- paste0("D:/newmove/", mouse_id, "/100-0/", session, "/trialsummary.txt")
    if (trial_data_key %in% names(trial_lookup)) {
      tracking_list[[i]] <- add_trial_label(tracking_list[[i]], trial_lookup[[trial_data_key]])
    } else {
      tracking_list[[i]]$trial_label <- "Between trials"
    }
  }
  tracking_list
}

doi_tracking_data_processed <- label_all_trial_periods(doi_tracking_data, trial_data_lookup)
saline_tracking_data_processed <- label_all_trial_periods(saline_tracking_data, trial_data_lookup)
```

# HTR event import and matching

```{r import htr timestamps}
# htr timestamps from two sources: automated detection ("D:/Time Stamps
# Task") and manually-scored files ("D:/Time Stamps Task Logan")
read_htr_timestamps <- function(file) {
  raw_data <- readBin(file, "raw", file.info(file)$size)
  if (length(raw_data) >= 2 && raw_data[1] == 0xff && raw_data[2] == 0xfe) {
    raw_data <- raw_data[-(1:2)]
  } else if (length(raw_data) >= 2 && raw_data[1] == 0xfe && raw_data[2] == 0xff) {
    raw_data <- raw_data[-(1:2)]
  }
  raw_data <- raw_data[raw_data != 0]
  text_data <- rawToChar(raw_data)
  tmp_file <- tempfile()
  writeLines(text_data, tmp_file)
  read.csv(tmp_file, header = FALSE, col.names = "Timestamp", stringsAsFactors = FALSE)
}

htr_list <- list.files(path = "D:/Time Stamps Task", pattern = "*.csv", full.names = TRUE)
htr_data <- lapply(htr_list, read_htr_timestamps)
names(htr_data) <- basename(htr_list)

# convert HH:MM:SS:MS timestamp strings to seconds
convert_to_seconds <- function(timestamp) {
  time_parts <- strsplit(timestamp, ":")[[1]]
  ms <- if (length(time_parts) == 4) as.numeric(strsplit(time_parts[4], "\\.")[[1]][1]) else 0
  as.numeric(time_parts[1]) * 3600 + as.numeric(time_parts[2]) * 60 + as.numeric(time_parts[3]) + ms / 1000
}

htr_seconds <- lapply(htr_data, function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col) && grepl(":", col[1])) return(sapply(col, convert_to_seconds))
    col
  })
  df
})

# manually-scored htr files (already in seconds)
logan_paths <- list.files("D:/Time Stamps Task Logan", full.names = TRUE)
logan_names <- basename(logan_paths)
logan_htr <- setNames(lapply(logan_paths, function(f) {
  df <- read.csv(f)
  colnames(df) <- "Timestamp"
  df
}), logan_names)

htr_seconds <- c(logan_htr, htr_seconds)
```

```{r mark htr frames}
# tags each tracking frame "yes"/"no" for whether an htr occurred on it
# (nearest tracking frame to each htr timestamp)
mark_htr_frames <- function(tracking_data, htr_data) {
  if (!"time" %in% colnames(tracking_data)) {
    tracking_data$time <- (1:nrow(tracking_data)) / 80
  }
  tracking_data <- tracking_data %>% mutate(htr = "no")
  closest_indices <- map_int(htr_data$Timestamp, function(htr_time) {
    which.min(abs(tracking_data$time - htr_time))
  })
  tracking_data$htr[closest_indices] <- "yes"
  tracking_data
}

processed_tracking_data <- list()
for (tracking_list in list(doi_tracking_data_processed, saline_tracking_data_processed)) {
  for (filename in names(tracking_list)) {
    mouse_id_session <- sub("_(doi|saline)\\.csv$", "", filename)
    lookup_name <- paste0(mouse_id_session, ".csv")
    if (!lookup_name %in% names(htr_seconds)) next

    htr_df <- htr_seconds[[lookup_name]]
    if (!is.data.frame(htr_df) || !"Timestamp" %in% colnames(htr_df)) next

    tracking_data <- tracking_list[[filename]]
    if (!is.data.frame(tracking_data) || nrow(tracking_data) == 0) next

    tryCatch({
      processed_tracking_data[[filename]] <- mark_htr_frames(tracking_data, htr_df)
    }, error = function(e) {
      message(paste("error processing", filename, ":", e$message))
    })
  }
}
```

```{r split and flip}
# split into doi/saline (only mice with matched htr timestamp files), and
# flip x-coordinates for the 5 mice recorded on the earlier camera setup
flip_x <- function(tracking_list) {
  lapply(names(tracking_list), function(name) {
    df <- tracking_list[[name]]
    if (gsub("_.*", "", name) %in% flip_ids) df$x <- 1200 - df$x
    df
  }) %>% setNames(names(tracking_list))
}

tracking_doi <- flip_x(processed_tracking_data[grepl("doi.csv$", names(processed_tracking_data))])
tracking_saline <- flip_x(processed_tracking_data[grepl("saline.csv$", names(processed_tracking_data))])

# htr-only frames (figure 5c), and full long-format frames (figure 5d
# occupancy-normalized htr map), both source-tagged by mouse_session
tracking_doi_combined <- bind_rows(lapply(tracking_doi, filter, htr == "yes"), .id = "source")
tracking_saline_combined <- bind_rows(lapply(tracking_saline, filter, htr == "yes"), .id = "source")

tracking_doi_long <- bind_rows(tracking_doi, .id = "source")
tracking_saline_long <- bind_rows(tracking_saline, .id = "source")
```

# Figure 5C - nose location at HTR onset

```{r fig 5c}
tracking_doi_combined %>%
  ggplot(aes(x = x, 
             y = y, 
             color = trial_label)) +
  geom_point(size = 2, 
             alpha = 0.7) +
  ggtitle("Nose location at HTR onset") +
  theme_classic() +
  scale_color_manual(values = c("Between trials" = "#30acea", 
                                "Within trial" = "#5f009c"),
                     labels = c("Between trials" = "ITI", 
                                "Within trial" = "Within trial")) +
  theme(axis.title  = element_blank(),
        axis.text   = element_blank(),
        axis.ticks  = element_blank(),
        axis.line   = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 20),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.position = "bottom")
```

# Figure 5D and occupancy-normalized HTR maps

```{r htr raw binned counts}
# figure 5d: raw htr counts per bin (~2.5 cm bins), pooled across all doi
# mice - not yet normalized by occupancy
bin_size <- 100  # pixels; ~2.5 cm given the 0.0237 cm/px conversion

tracking_doi_combined_nonzero <- tracking_doi_combined %>%
  mutate(x = ifelse(x < 0, 0, x), y = ifelse(y < 0, 0, y),
         x_bin = floor(x / bin_size) * bin_size,
         y_bin = floor(y / bin_size) * bin_size)

htr_density_nonzero_filtered <- tracking_doi_combined_nonzero %>%
  group_by(x_bin, y_bin) %>%
  summarise(htr_count = sum(htr == "yes"), .groups = "drop") %>%
  filter(x_bin != 1200, y_bin != 700)  # exclude bins right at the arena edge

write.csv(htr_density_nonzero_filtered, "D:/R data/htr_binned.csv", row.names = FALSE)
```

```{r htr normalized by occupancy per mouse}
# occupancy-normalize htr rate (htr/min) per mouse per bin, then average
# across mice - separately for all frames, within-trial, and iti
bin_width <- 100

tracking_doi_long_nonzero <- tracking_doi_long %>%
  filter(speed_aw_cm < 160) %>%
  mutate(x = ifelse(x < 0, 0, x), y = ifelse(y < 0, 0, y))

master_bins <- tracking_doi_long_nonzero %>%
  mutate(bin_x = floor(x / bin_width) * bin_width, bin_y = floor(y / bin_width) * bin_width) %>%
  distinct(source, bin_x, bin_y)

compute_occupancy_by_mouse <- function(data, trial_filter = NULL) {
  if (!is.null(trial_filter)) data <- data %>% filter(trial_label == trial_filter)
  data %>%
    mutate(bin_x = floor(x / bin_width) * bin_width, bin_y = floor(y / bin_width) * bin_width) %>%
    group_by(source, bin_x, bin_y) %>%
    summarise(occupancy_time_seconds = n() / 80, .groups = "drop")
}

compute_htr_map_by_mouse <- function(data, trial_filter = NULL) {
  if (!is.null(trial_filter)) data <- data %>% filter(trial_label == trial_filter)
  data %>%
    filter(htr == "yes") %>%
    mutate(bin_x = floor(x / bin_width) * bin_width, bin_y = floor(y / bin_width) * bin_width) %>%
    group_by(source, bin_x, bin_y) %>%
    summarise(htr_count = n(), .groups = "drop")
}

occupancy_by_mouse <- compute_occupancy_by_mouse(tracking_doi_long_nonzero)
occupancy_by_mouse_within <- compute_occupancy_by_mouse(tracking_doi_long_nonzero, "Within trial")
occupancy_by_mouse_iti <- compute_occupancy_by_mouse(tracking_doi_long_nonzero, "Between trials")

htr_map_by_mouse <- compute_htr_map_by_mouse(tracking_doi_long_nonzero)
htr_map_by_mouse_within <- compute_htr_map_by_mouse(tracking_doi_long_nonzero, "Within trial")
htr_map_by_mouse_iti <- compute_htr_map_by_mouse(tracking_doi_long_nonzero, "Between trials")

htr_normalized_permouse <- full_join(htr_map_by_mouse, occupancy_by_mouse, by = c("source", "bin_x", "bin_y")) %>%
  mutate(htr_rate = (htr_count / occupancy_time_seconds) * 60) %>%
  replace_na(list(htr_rate = 0))

htr_normalized_permouse_iti <- full_join(htr_map_by_mouse_iti, occupancy_by_mouse_iti, by = c("source", "bin_x", "bin_y")) %>%
  mutate(htr_rate = (htr_count / occupancy_time_seconds) * 60) %>%
  replace_na(list(htr_rate = 0))

# within-trial bins are 0-filled against the master grid first (some mice
# have zero within-trial occupancy in bins other mice do visit)
occupancy_by_mouse_within_filled <- master_bins %>%
  left_join(occupancy_by_mouse_within, by = c("source", "bin_x", "bin_y")) %>%
  replace_na(list(occupancy_time_seconds = 0))
htr_map_by_mouse_within_filled <- master_bins %>%
  left_join(htr_map_by_mouse_within, by = c("source", "bin_x", "bin_y")) %>%
  replace_na(list(htr_count = 0))
htr_normalized_permouse_within <- occupancy_by_mouse_within_filled %>%
  left_join(htr_map_by_mouse_within_filled, by = c("source", "bin_x", "bin_y")) %>%
  mutate(htr_rate = ifelse(occupancy_time_seconds > 0, (htr_count / occupancy_time_seconds) * 60, 0))
```

```{r grand mean htr maps}
grand_mean_htr <- htr_normalized_permouse %>%
  group_by(bin_x, bin_y) %>%
  summarise(mean_htr_rate = mean(htr_rate, na.rm = TRUE), .groups = "drop") %>%
  filter(bin_y != 700, bin_x != 1200)

grand_htr_within <- htr_normalized_permouse_within %>%
  group_by(bin_x, bin_y) %>%
  summarise(mean_htr_rate = mean(htr_rate, na.rm = TRUE), .groups = "drop") %>%
  filter(bin_y != 700, bin_x != 1200)

grand_htr_iti <- htr_normalized_permouse_iti %>%
  group_by(bin_x, bin_y) %>%
  summarise(mean_htr_rate = mean(htr_rate, na.rm = TRUE), .groups = "drop") %>%
  filter(bin_y != 700, bin_x != 1200)

write.csv(grand_mean_htr, "D:/R data/htr_grandmean.csv", row.names = FALSE)
write.csv(grand_htr_iti, "D:/R data/htr_grandmean_iti.csv", row.names = FALSE)
write.csv(grand_htr_within, "D:/R data/htr_grandmean_within.csv", row.names = FALSE)
```

# Figure supplemental - spatial permutation tests

```{r permutation test x y coordinates}
# tests whether htr rate correlates with bin position (x and y
# separately), by comparing the observed correlation to a null
# distribution built by shuffling bin position 10,000 times.
run_position_permutation_test <- function(data, n_permutations = 10000, seed = 123) {
  set.seed(seed)
  observed_corr_x <- cor(data$htr_rate, data$bin_x)
  observed_corr_y <- cor(data$htr_rate, data$bin_y)

  permuted <- matrix(NA, nrow = n_permutations, ncol = 2)
  for (i in 1:n_permutations) {
    permuted[i, 1] <- cor(data$htr_rate, sample(data$bin_x))
    permuted[i, 2] <- cor(data$htr_rate, sample(data$bin_y))
  }

  list(
    p_value_x = mean(abs(permuted[, 1]) >= abs(observed_corr_x)),
    p_value_y = mean(abs(permuted[, 2]) >= abs(observed_corr_y)),
    observed_corr_x = observed_corr_x,
    observed_corr_y = observed_corr_y
  )
}

htr_normalized_permouse_filtered <- htr_normalized_permouse %>% filter(bin_y != 700, bin_x != 1200)
htr_normalized_permouse_within_filtered <- htr_normalized_permouse_within %>% filter(bin_y != 700, bin_x != 1200)

# combined (within-trial + iti pooled) - figure 5e
perm_test_combined <- run_position_permutation_test(htr_normalized_permouse_filtered)
print("permutation test, combined (all trial periods):")
print(perm_test_combined)

# within-trial only - figure 4 supplemental
perm_test_within <- run_position_permutation_test(htr_normalized_permouse_within_filtered)
print("permutation test, within trial only:")
print(perm_test_within)
```

# Figure 5F - HTR rate, within trial vs. ITI (+ sex-difference supplemental)

this uses a separate, simpler pipeline (raw htr timestamps + trial
timing lookup) rather than the tracking-based pipeline above.

```{r htr rate by trial period}
doi_conditions <- paste(mouse_sessions$mouse_id, mouse_sessions$session_doi, sep = "_")

get_id_session <- function(name) str_extract(basename(name), "\\d+_\\d+")

htr_rates <- list()
for (i in seq_along(htr_seconds)) {
  htr_df <- htr_seconds[[i]]
  id_session <- get_id_session(names(htr_seconds)[i])
  id <- str_split(id_session, "_")[[1]][1]
  session <- str_split(id_session, "_")[[1]][2]

  match_index <- which(str_detect(names(trial_data_lookup), paste0("/", id, "/.*?/", session, "/")))
  if (length(match_index) != 1) next

  trial_df <- trial_data_lookup[[match_index]]
  tstart <- trial_df$tstart
  tend <- trial_df$tend

  htr_df <- htr_df %>% mutate(within_trial = map_lgl(Timestamp, function(ts) any(ts >= tstart & ts <= tend)))

  total_trial_time <- sum(tend - tstart, na.rm = TRUE)
  total_time <- 2400  # 40 min session, in seconds
  iti_time <- total_time - total_trial_time

  htrs_in_trial <- sum(htr_df$within_trial)
  htrs_in_iti <- nrow(htr_df) - htrs_in_trial

  htr_rates[[id_session]] <- data.frame(
    id = as.numeric(id), session = as.numeric(session),
    htrs_in_trial = htrs_in_trial, htrs_in_iti = htrs_in_iti,
    total_trial_time = total_trial_time, iti_time = iti_time,
    rate_in_trial = htrs_in_trial / total_trial_time,
    rate_in_iti = htrs_in_iti / iti_time
  )
}

htr_rates_doi <- htr_rates[names(htr_rates) %in% doi_conditions]
htr_rate_df_doi <- bind_rows(htr_rates_doi, .id = "id_session") %>%
  mutate(
    rate_in_trial_per_min = rate_in_trial * 60,
    rate_in_iti_per_min = rate_in_iti * 60,
    sex = ifelse(id %in% female_ids, "F", "M"),
    htr = htrs_in_trial + htrs_in_iti,
    htr_rate = htr / 40  # htr per minute over the full 40-min session
  )

group_stats_by_sex <- htr_rate_df_doi %>%
  group_by(sex) %>%
  summarize(iti_median = median(rate_in_iti_per_min, na.rm = TRUE),
            trial_median = median(rate_in_trial_per_min, na.rm = TRUE),
            n = n())
```

```{r fig 5f}
htr_rate_df_doi %>%
  ggplot(aes(x = rate_in_trial_per_min, y = rate_in_iti_per_min, color = sex, shape = sex)) +
  geom_point(size = 6, stroke = 1.5, alpha = 0.8) +
  geom_point(data = group_stats_by_sex,
             aes(x = trial_median, y = iti_median, color = sex),
             size = 10, shape = 1, stroke = 2.5, inherit.aes = FALSE) +
  scale_color_manual(values = c("M" = "#00B19E", "F" = "#990068"), guide = "none") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black", size = 1) +
  labs(title = "HTR rate", x = "Within trial (HTR/min)", y = "ITI (HTR/min)", shape = "Sex") +
  scale_x_continuous(limits = c(0, 6.5)) +
  scale_y_continuous(limits = c(0, 6.5)) +
  coord_fixed(ratio = 1) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.15),
        plot.title = element_text(hjust = 0.5, size = 24),
        legend.title = element_blank(),
        text = element_text(size = 20),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2)) +
  guides(shape = guide_legend(override.aes = list(color = c("#990068", "#00B19E"))))
```

```{r fig 5f stats}
wilcox_5f <- wilcox.test(htr_rate_df_doi$rate_in_trial_per_min, htr_rate_df_doi$rate_in_iti_per_min, paired = TRUE)
print(wilcox_5f)
```

```{r fig supplemental htr rate by sex}
ggplot(htr_rate_df_doi, aes(x = sex, y = htr_rate, color = sex)) +
  geom_boxplot(aes(fill = sex), alpha = 0.2, color = "black", outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 17, size = 5, color = "red") +
  geom_beeswarm(size = 3, stroke = 1.5, alpha = 0.8, cex = 2, jitter = 1) +
  geom_line(aes(group = sex), color = "gray", alpha = 0.8) +
  labs(x = NULL, y = "HTR/min", color = "Sex", title = "HTR rate") +
  scale_fill_manual(values = c("F" = "#990068", "M" = "#00899c")) +
  scale_color_manual(values = c("F" = "#990068", "M" = "#00899c")) +
  scale_x_discrete(labels = function(x) gsub(" ", "\n", x)) +
  theme_classic() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, size = 24),
        text = element_text(size = 20),
        legend.text = element_text(size = 20),
        axis.title = element_text(size = 20),
        axis.text = element_text(size = 20),
        axis.line = element_line(linewidth = 1.2))
```

```{r fig 4 supplemental sex difference}
wilcox_htr_sex <- wilcox.test(htr_rate ~ sex, data = htr_rate_df_doi, exact = FALSE)
print(wilcox_htr_sex)
```

# Figure 5A/5B - occupancy, DOI vs. saline data for MATLAB plotting

all 22 mice (not just the subset with matched htr timestamps), ~1 cm
bins (matching Reese et al.'s bin size), speed-filtered and x-flipped
the same way as above.

```{r occupancy all 22 mice}
bin_width_occupancy <- 48  # ~1 cm^2 bins

tracking_doi_long_all22 <- bind_rows(doi_tracking_data_processed, .id = "source") %>%
  mutate(source_prefix = gsub("_.*", "", source), x = ifelse(source_prefix %in% flip_ids, 1200 - x, x)) %>%
  select(-source_prefix) %>%
  filter(speed_aw_cm < 160)

tracking_saline_long_all22 <- bind_rows(saline_tracking_data_processed, .id = "source") %>%
  mutate(source_prefix = gsub("_.*", "", source), x = ifelse(source_prefix %in% flip_ids, 1200 - x, x)) %>%
  select(-source_prefix) %>%
  filter(speed_aw_cm < 160)

compute_occupancy_22 <- function(data, trial_filter = NULL) {
  if (!is.null(trial_filter)) data <- data %>% filter(trial_label == trial_filter)
  data %>%
    mutate(bin_x = floor(x / bin_width_occupancy) * bin_width_occupancy,
           bin_y = floor(y / bin_width_occupancy) * bin_width_occupancy) %>%
    group_by(source, bin_x, bin_y) %>%
    summarise(occupancy_time_seconds = n() / 80, .groups = "drop")
}

occupancy_by_mouse_22_doi <- compute_occupancy_22(tracking_doi_long_all22)
occupancy_22_doi_within_trial <- compute_occupancy_22(tracking_doi_long_all22, "Within trial")
occupancy_22_doi_iti <- compute_occupancy_22(tracking_doi_long_all22, "Between trials")

occupancy_by_mouse_22_saline <- compute_occupancy_22(tracking_saline_long_all22)
occupancy_22_saline_within_trial <- compute_occupancy_22(tracking_saline_long_all22, "Within trial")
occupancy_22_saline_iti <- compute_occupancy_22(tracking_saline_long_all22, "Between trials")

grand_mean_across_mice <- function(data, value_col = "occupancy_time_seconds", out_col = "mean_occupancy") {
  data %>%
    group_by(bin_x, bin_y) %>%
    summarise(!!out_col := mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
}

occupancy_all_doi <- grand_mean_across_mice(occupancy_by_mouse_22_doi)
occupancy_all_doi_within_trial <- grand_mean_across_mice(occupancy_22_doi_within_trial)
occupancy_all_doi_iti <- grand_mean_across_mice(occupancy_22_doi_iti)

occupancy_all_saline <- grand_mean_across_mice(occupancy_by_mouse_22_saline)
occupancy_all_saline_within_trial <- grand_mean_across_mice(occupancy_22_saline_within_trial)
occupancy_all_saline_iti <- grand_mean_across_mice(occupancy_22_saline_iti)

# fraction of a session's frames spent in each bin (fixes the broken
# export noted at the top of this file)
total_frames_per_source <- tracking_doi_long_all22 %>%
  group_by(source) %>%
  summarise(total_frames_in_source = n(), .groups = "drop")

occupancy_doi_frames <- tracking_doi_long_all22 %>%
  mutate(bin_x = floor(x / bin_width_occupancy) * bin_width_occupancy,
         bin_y = floor(y / bin_width_occupancy) * bin_width_occupancy) %>%
  group_by(source, bin_x, bin_y) %>%
  summarise(occupancy_time_frames = n(), .groups = "drop") %>%
  left_join(total_frames_per_source, by = "source") %>%
  mutate(fraction_of_total_frames = occupancy_time_frames / total_frames_in_source) %>%
  group_by(bin_x, bin_y) %>%
  summarise(mean_frames = mean(fraction_of_total_frames, na.rm = TRUE), .groups = "drop")
```

```{r export occupancy csvs}
# uncomment code to save files to import into MATLAB for plotting

# write.csv(occupancy_doi_frames, "D:/R data/occupancy_doi_frames.csv", row.names = FALSE)
# 
# write.csv(occupancy_all_doi, "D:/R data/occupancy_doi_time.csv", row.names = FALSE)
# write.csv(occupancy_all_doi_within_trial, "D:/R data/occupancy_doi_within.csv", row.names = FALSE)
# write.csv(occupancy_all_doi_iti, "D:/R data/occupancy_doi_iti.csv", row.names = FALSE)
# 
# write.csv(occupancy_all_saline, "D:/R data/occupancy_saline_time.csv", row.names = FALSE)
# write.csv(occupancy_all_saline_within_trial, "D:/R data/occupancy_saline_within.csv", row.names = FALSE)
# write.csv(occupancy_all_saline_iti, "D:/R data/occupancy_saline_iti.csv", row.names = FALSE)
```
